---
title: "PMIM102 Resubmission Part 2"
author: "2140714"
date: '2022-12-02'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

- Description of Health Boards in Wales to choose from in Wales

7A1: Betsi Cadwaladr University Health Board
7A2: Hywel Dda University Health Board
7A3: Abertawe Bro Morgannwg University Local Health Board
7A4: Cardiff and Vale University Health Board
7A5: Cwm Taf Morgannwg University Health Board
7A6: Aneurin Bevan Local Health Board
7A7: Powys Teaching Local Health Board

- Objectives:
1.0: What is the total prescriptions from April 2013 to December 2015?
2.0: Find out information about the prescription by health boards in Wales
3.0: What is the total items prescribed for therapeutic classes
4.0: Analyse and visualise the number of items prescribed/percentage time change from the top 2 largest therapeutic classes ('Cardiovascular System' and 'Endrocrine System') between April 2013 and December 2015.
5.0: Find the rate of diabetes in all the different local Health Boards in Wales
6.0: Find the statistical relationship between the rate of dementia vs the rate of smoking and diabetes


```{r}
#install these packages if you do not have them
#install.packages("tidyverse")
#install.packages(RPostgreSQL)
#install.packages(GetoptLong) 
#install.packages(ggplot2)

# Load the required packages
library(tidyverse)
library(RPostgreSQL)
library(GetoptLong)
library(ggplot2)

theme_set(theme_minimal()) #preset charts to have a minimalistic theme
```

```{r}
# connect R to PostgreSQL to access the database and run SQL
tryCatch({ 
drv <- dbDriver("PostgreSQL")
print("Connecting to Database…")
con <- dbConnect(drv, dbname = "gp_practice_data", # edit this if your database is different in PostgreSQL
                 host = "localhost", 
                 port = 5432, # edit this if your port is different in PostgreSQL
                 user = "postgres", 
                 password = rstudioapi::askForPassword())
 print("Database Successfully Connected!")
    },
    error=function(con) {
            print("Unable to connect to Database.")
    })
```

```{r}
# validate the data from the database
# check for the available tables in the database
cat('\nBelow are the following available tables gotten from the Database:\n')
tables <- dbListTables(con)
print(tables)
```
```{r}
# Get the list of Health Boards to choose from in Wales
hb_list <- dbGetQuery(con, qq("select practiceid as hb_id, locality, street, area, posttown, county, postcode
                              from address
                              WHERE practiceid like '7%'
                              order by practiceid asc;"))

cat('\nBelow is a table containing information on Health Boards in Wales along with their Unique code (hb). There are', nrow(hb_list), 'health boards in Wales.\n')
print(hb_list)
```

We would perform some analysis using the local health boards
- Allow user to select a Health Board using the unique ID
```{r}
# Data validation
# Allow user to select a health board. 
# Health boards are represented by a unique code between 7A1 and 7A7.
User_Input <- function(){ 
  a <- ""
  while(a == ""){
    a <- readline(prompt="Enter Health Board (7Ax): ")
    if(tolower(a) == "q"){
      break
    } 
    if(grepl('^7A[1-7]{1}$', a)==FALSE){
      print ("Data does not exist")
      print("Please try again or select another Health Board.")
      a <- "" 
    } 
  } 
  return(a)
}

Health_Board <- User_Input()
cat('\nDone!.. Health Board is Available\n\n')
```

```{r}
# Get health board from user input
# Check if the user selected Health Board has prescription data available
User_Hb <- dbGetQuery(con, qq('select *
                              from gp_data_up_to_2015
                              where hb = \'@{Health_Board}\''))
                              
cat('\nLoading information, make take some time! please wait...\n\n')
print(User_Hb)
```


# Objective 1: What is the total prescriptions from April 2013 to December 2015?

# 1.1
List of prescribed items in the selected health board from April 2013 to December 2015
```{r}
hb_pres_list <- dbGetQuery(con, qq('select hb, period, bnfcode, items
                            from gp_data_up_to_2015
                            where hb = \'@{Health_Board}\'
                            group by hb, bnfcode, items, period')) 

# Group by sum
hb_pres_tbl <- hb_pres_list %>% group_by(period, hb) %>% 
              summarise(total_pres = sum(items),
            .groups = 'drop') %>% as.data.frame()

# recode the health board
hb_pres_tbl$hb <- recode(hb_pres_tbl$hb,
                           "7A1" = "Betsi Cadwaladr University Health Board",
                           "7A2" = "Hywel Dda University Health Board",
                           "7A3" = "Abertawe Bro Morgannwg University Local Health Board",
                           "7A4" = "Cardiff and Vale University Health Board",
                           "7A5" = "Cwm Taf Morgannwg University Health Board",
                           "7A6" = "Aneurin Bevan Local Health Board",
                           "7A7" = "Powys Teaching Local Health Board")

cat('The table contains the total items prescribed in', Health_Board, 'between 2013/04 and 2015/12:\n', sep=' ')
print(hb_pres_tbl)

```


# 1.2

The total monthly spend on prescription in the selected health board from April 2013 to December 2015
```{r}
## plot the monthly trend from 2013 to 2015
# convert period to date format
hb_pres_tbl$date <- format(as.Date(paste(hb_pres_tbl$period, '01'), '%Y%m%d'), '%Y-%m-%d')
hb_pres_tbl$date <- as.Date(hb_pres_tbl$date)
str(hb_pres_tbl) # confirm the structure

# Plot chart
plot1 <- ggplot(hb_pres_tbl, aes(y=total_pres, x=date, fill = ifelse(total_pres == min(total_pres), "Lowest", ifelse(total_pres == max(total_pres), "Highest", "Others")))) + 
  geom_bar(position="dodge", stat="identity") +
  ggtitle("Monthly Total of items prescribed in the selected Health Board between April 2013 and December 2015") +
  ylab("") + xlab("") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(name = "Total Prescriptions", values = c("Lowest" = "#1B998B", "Highest" = "#287271", "Others" = "steelblue")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())



cat('\nPlotting monthly total of items prescribed in', Health_Board, 'between April 2013 and December 2015:\n')
print(plot1)

```


Objective 2.0: Find out information about the prescription for different classes and in health boards

# 2.1

From the data 'bnfchapter' is the unique code that represents the disease treatment/therapeutic class.

The unique code (bnfchapter) are digits between 01 and 15, 18 and 23, each representing different treatment/therapeutic class. We can furthermore see that these digits are the first two digits in' bnfchemical'. With this information we can deduce the treatment/therapeutic category from prescription data.

```{r}
# extract therapeutic class data   
dis_cat <- dbGetQuery(con, qq('select distinct bnfchapter, chapterdesc
                              from bnf
                              order by bnfchapter asc'))

cat('\nLoading therapeutic class, please wait...\n\n')
print(dis_cat)
```
- The table above show the code representing the body's therapeutic system


# 2.2

# Create a new variable (med_id) from the first two digits of the bnfcode from 'gp_data_up_to_2015' data
# This variable would help us identify different therapeutic classes as discovered in bnf data
```{r}
## Create a variable with the first two digits from the bnfcode. This will help us identify
## the body system that each item was prescribed for.

dis_tbl <- dbGetQuery(con, qq('SELECT LEFT(bnfcode, 2) AS med_id, 
                              hb, items, nic, actcost, period
                              FROM gp_data_up_to_2015'))

cat('\nLoading information, make take some time! please wait...\n\n')
print(dis_tbl)

# validate the newly created table
no_row <- nrow(dis_tbl) 
cat('\nThe total amount of entries this new table has is:', no_row,'. The new table is validated as it still contains 21596397 entries corresponding with the original gp_data_up_to_2015 data. This indicates that after creating a new variable the data has not been distorted.\n\n')
```

# 2.3

Q: Find the total number items prescribed to treat different therapeutic classes in health boards in Wales
```{r}
## select the required variables needed to find the total number of items prescribed for the different therapeutic classes
class_treat_all <- dis_tbl %>% select(med_id, hb, items, period)

## recode the therapeutic classes from the code to their description
class_treat_all$med_id <- recode(class_treat_all$med_id,
                           "01" = "Gastro-Intestinal System",
                           "02" = "Cardiovascular System",
                           "03" = "Respiratory System",
                           "04" = "Central Nervous System",
                           "05" = "Infections",
                           "06" = "Endocrine System",
                           "07" = "Obstetrics Gynae+Urinary Tract Disorders",
                           "08" = "Malignant Disease & Immunosuppression",
                           "09" = "Nutrition And Blood",
                           "10" = "Musculoskeletal & Joint Diseases",
                           "11" = "Eye",
                           "12" = "Ear, Nose And Oropharynx",
                           "13" = "Skin",
                           "14" = "Immunological Products & Vaccines",
                           "15" = "Anaesthesia",
                           "18" = "Preparations used in Diagnosis",
                           "19" = "Other Drugs And Preparations",
                           "20" = "Dressings",
                           "21" = "Appliances",
                           "22" = "Incontinence Appliances",
                           "23" = "Stoma Appliances")

# recode the local health boards from their code to the names of the health board
class_treat_all$hb <- recode(class_treat_all$hb,
                           "7A1" = "Betsi Cadwaladr University Health Board",
                           "7A2" = "Hywel Dda University Health Board",
                           "7A3" = "Abertawe Bro Morgannwg University Local Health Board",
                           "7A4" = "Cardiff and Vale University Health Board",
                           "7A5" = "Cwm Taf Morgannwg University Health Board",
                           "7A6" = "Aneurin Bevan Local Health Board",
                           "7A7" = "Powys Teaching Local Health Board")


cat('\nThe table shows the total items prescribed in the local health boards\n\n')
print(class_treat_all)
```
# 2.4

# Visualisation: 'Prescription in Health Boards'
# Plot the total number of items prescribed in the Health Boards in Wales between April 2013 and December 2015
```{r}
# Analyse the total number of items prescribed by Health Boards in Wales between April 2013 and December 2015
class_tbl_all <- class_treat_all %>%
  group_by(med_id, hb, period) %>%
  # Calculate the total value for each class and order from top to bottom
  summarize(total = sum(items)) %>% ungroup() 

# set the items variable to numeric
class_tbl_all$total <- as.numeric(class_tbl_all$total)

hb_plot2 <- ggplot(class_tbl_all, aes(x = total, y = reorder(hb, total))) +
                  ggtitle("Total number of items prescribed by Health Boards in Wales (April                   2013 to December 2015") + 
                  ylab("") + xlab("") +
                  geom_bar(fill = "#287271", stat = "identity") +
                  scale_x_continuous(labels = scales::comma) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())


print(hb_plot2)
cat('\nPlotting, please wait...\n\n')
cat('\nPowys Teaching Health Board prescribed fewer items than any other Health Boards in Wales\n\n')
```


# Objective 3.0: What is the total items prescribed for therapeutic classes

# 3.1
# Visualisation: Prescription for different therapeutic classes
```{r}
# plot total items prescribed by therapeutic class in health boards
class_plot3 <- ggplot(class_tbl_all, aes(y = reorder(med_id, total), x = total)) + 
                  geom_bar(fill = "#287271", stat = "identity") +
                  ggtitle("Total Number of Items Prescribed by classes 
                  between April 2013 and December 2015 in Wales") + 
                  ylab("") + xlab("") +
                  labs(fill = "Local Health Boards") +
                  scale_x_continuous(labels = scales::comma) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

cat('\nPlotting total of items prescribed for therapeutic classes in local health boards in Wales\n\n')
print(class_plot3)
cat('\nThere are no items prescribed in the ‘preparations used in diagnosis’ class between April 2015 and December 2015\n\n')
cat('\nThe top two therapeutic classes that had the highest number of prescriptions were Cardiovascular System and Central Nervous System \n\n')

```


# Objective 4.0: Analyse and visualise the number of items prescribed/percentage time change from the top 2 largest therapeutic classes ('Cardiovascular System' and 'Endrocrine System') between April 2013 and December 2015.

# 4.1
# Analysis for Class: Cardiovascular system
```{r}
# Cardiovascular analysis
# select thevariables needed for this analysis
test_tbl <- class_treat_all %>% select(med_id, items, period)
head(test_tbl) # preveiew the data


# filter the data by the class we intend to analyse: 'Cardiovascular System'
cardio_class <- test_tbl %>% filter(med_id == "Cardiovascular System")
head(cardio_class) # preview the data

# create correct date format
# Add the day "01" to the date column
cardio_class$date <- format(as.Date(paste(cardio_class$period, '01'), '%Y%m%d'), '%Y-%m-%d')
cardio_class$date <- as.Date(cardio_class$date)

# validate the date variable is properly format by looking at its structure
str(cardio_class)

# sum the total items prescribed and group by the date and class and create a new table
cardio_tbl <- cardio_class %>%
  group_by(date, med_id) %>%
  # Calculate the total value for each group
  summarize(total = sum(items)) %>% ungroup()

head(cardio_tbl) # preview the data

# get percentage
cardio_tbl$percent <- cardio_tbl$total / sum(cardio_tbl$total)

# Create the plot
cardio_plot4 <- ggplot(cardio_tbl, aes(x = date, y = percent)) +
                      geom_line() +
  ggtitle("Yearly change of items prescibed for Cardiovascular Class between April 2013 and December 2015") +
                      labs(x = "Date", y = "Number of items prescribed") +
                      ylab ("") + xlab ("") +
                      scale_y_continuous(labels = scales::percent) +
                      theme(panel.grid.major = element_blank(), 
                      panel.grid.minor = element_blank())

print(cardio_plot4)
cat('\nPlotting, please wait...\n\n')
```

# 4.2
# Analysis for Class: Central Nervous System
```{r}
#Central Nervous System analysis

# filter the data by the class we intend to analyse: 'Endrocrine System'
cns_class <- test_tbl %>% filter(med_id == "Central Nervous System")
head(cns_class) # preview the data

# create correct date format
# Add the day "01" to the date column
cns_class$date <- format(as.Date(paste(cns_class$period, '01'), '%Y%m%d'), '%Y-%m-%d')
cns_class$date <- as.Date(cns_class$date)

# validate the date variable is properly format by looking at its structure
str(cns_class)

# sum the total items prescribed and group by the date and class and create a new table
cns_tbl <- cns_class %>%
  group_by(date, med_id) %>%
  # Calculate the total value for each group
  summarize(total = sum(items)) %>% ungroup()

head(cns_tbl) # preview the data

# get percentage
cns_tbl$percent <- cns_tbl$total / sum(cns_tbl$total)

# Create the plot
cns_plot5 <- ggplot(cns_tbl, aes(x = date, y = percent)) +
                    geom_line() +
                    ggtitle("Yearly change of total items prescribed for Central Nervous System between April 2013 and December 2015") +
                    labs(x = "Date", y = "Number of items prescribed") +
                    ylab ("") + xlab("") +
                    scale_y_continuous(labels = scales::percent) +
                    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

print(cns_plot5)
cat('\nPlotting, please wait...\n\n')
```



# Objective 5.0: Find the rate of diabetes in all the different local Health Boards in Wales

# 5.1 

# Analysis of the rate of diabetes

```{r}
# create a new table by joining disease information from qof_achievement to hb information from gp_data by practiceid
hb_dm <- dbGetQuery(con, qq("select gp_data_up_to_2015.hb, 
                          gp_data_up_to_2015.practiceid, 
                          qof_achievement.numerator, qof_achievement.field4, 
                          qof_achievement.indicator
                          from gp_data_up_to_2015
                          inner join qof_achievement on qof_achievement.orgcode =                                        gp_data_up_to_2015.practiceid
                          where qof_achievement.indicator like '%DM001%'
                          group by gp_data_up_to_2015.hb, 
                          gp_data_up_to_2015.practiceid, 
                          qof_achievement.numerator,
                          qof_achievement.field4,
                          qof_achievement.indicator;")) 

cat('\nLoading... please wait\n\n')
# recode the local health boards
hb_dm$hb <- recode(hb_dm$hb,
                           "7A1" = "Betsi Cadwaladr University Health Board",
                           "7A2" = "Hywel Dda University Health Board",
                           "7A3" = "Abertawe Bro Morgannwg University Local Health Board",
                           "7A4" = "Cardiff and Vale University Health Board",
                           "7A5" = "Cwm Taf Morgannwg University Health Board",
                           "7A6" = "Aneurin Bevan Local Health Board",
                           "7A7" = "Powys Teaching Local Health Board")

# Sum the numerator and total population
hba_tbl <- hb_dm %>% group_by(hb, indicator) %>% 
  summarise(numerator = sum(numerator), population = sum(field4),
            .groups = 'drop') %>% as.data.frame()


# get the rate of diabetes for each health board by diving the people with diabetes over the entire population
hba_tbl$rate <- hba_tbl$numerator / hba_tbl$population
cat('\nThe table shows the rate of diabetes in all the different local health boards\n\n')
print(hba_tbl)
```


# 5.2
```{r}
# plot a bar graph of the rate of diabetes in all the different health boards in Wales.
hb_plot6 <- ggplot(hba_tbl, aes(fill=hb, y=rate, x=indicator)) + 
                            geom_bar(position="dodge", stat="identity") +
                            ggtitle("Rate of Diabetes at Health Boards in Wales in 2015") + 
                            ylab("") + xlab("Rate of Diabetes") + 
                            labs(fill = "Local Health Boards in Wales") +
                            scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
                            scale_fill_manual(values=c("#003f5c", "#374c80", "#7a5195", "#bc5090", "#ef5675", "#ff764a", "#ffa600")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

cat('\nPlotting chart, please wait...\n\n')
print(hb_plot6)
```

Objective 6.0: Find the statistical relationship between the rate of dementia vs the rate of smoking and diabetes

# 6.1
```{r}
# Smoking and diabetes are risk factors for dementia. This section analyses if there is a relationship between the rate of
# diabetes, and smoking vs the rate of dementia in all the practices in Wales.

# rate of dementia in all the practices in Wales
dem_rate2 <- dbGetQuery(con, qq("select orgcode, ratio as rate_dem
                                from qof_achievement
                                where indicator like 'DEM001%' 
                                and orgcode not like 'WAL%'"))


# rate of diabetes in all the practices in Wales
dia_rate <- dbGetQuery(con, qq("select orgcode, ratio as rate_dia
                                from qof_achievement
                                where indicator like 'DM001%' 
                                and orgcode not like 'WAL%'"))

# rate of smoking in all the practices in Wales
smo_rate2 <- dbGetQuery(con, qq("select orgcode, ratio as rate_smo
                                from qof_achievement
                                where indicator like 'SMO SCR%' 
                                and orgcode not like 'WAL%'"))

# merge all tables (rate of dementia, diabetes and moking) by orgcode

#put all tables into list
rate_list <- list(dem_rate2, dia_rate, smo_rate2)      

#merge all tables together
cat('\nThe table below contains the rates of dementia, diabetes, smoking in all the practices in Wales.\n\n')
rate_tbl2 <- rate_list %>% reduce(full_join, by='orgcode') %>% as.data.frame()
print(rate_tbl2)

```

# 6.2

```{r}
# Rate of Dementia vs Diabetes in all the practices in Wales
Dem_dia <- plot(rate_tbl2$rate_dem ~ rate_tbl2$rate_dia,
     main = "Rate of Dementia vs Rate of Diabetes in all practices in Wales", ylab = "Rate of Dementia", 
     xlab = "Rate of Diabetes",
      pch = 19, col = "black")

cat('\nPlotting, please wait..\n\n')
print(Dem_dia)
```
```{r}
# Rate of Dementia vs Smoking in all the practices in Wales
Dem_smo <- plot(rate_tbl2$rate_dem ~ rate_tbl2$rate_smo,
     main = "Rate of Dementia vs Rate of Diabetes in all practices in Wales", ylab = "Rate of Dementia", 
     xlab = "Rate of Smoking",
      pch = 19, col = "black")

cat('\nPlotting, please wait..\n\n')
print(Dem_smo)
```

# Perform Correlation

# Correlation between the rate of dementia and the rate of diabetes
```{r}
cor_dem_dia <- cor.test(rate_tbl2$rate_dem, rate_tbl2$rate_dia)
cat('\nLoading correlation test, please wait..\n\n')
print(cor_dem_dia)

print(paste("The coefficient of the rate of dementia and diabetes is:", cor_dem_dia$estimate,"that indicates a weak positive correlation and with a p-value of:", cor_dem_dia$p.value,"that is below the significance level of 0.05"))
```
```{r}
cor_dem_smo <- cor.test(rate_tbl2$rate_dem, rate_tbl2$rate_smo)
cat('\nLoading correlation test, please wait..\n\n')
print(cor_dem_smo)

print(paste("The coefficient of the rate of dementia and smoking is:", cor_dem_smo$estimate,"that indicates a weak positive correlation and with a p-value of:", cor_dem_smo$p.value,"that is below the significance level of 0.05"))
```

```{r}
# Close the connection and unload the drivers.
dbDisconnect(con)
```

```{r}
cat('\nEnd of analysis. Thank you for using Harrison\'s code.\n',
    'For support, please contact Harrison on\n',
    'Imoetinharrison@gmail.com')
```
